Re-analysing the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}): A textbook illustration of the absence of evidence fallacy

### 2020-10-04 manuscript.tex
minor updates in manuscript

\abstract{
Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) reported the results of an experiment (N = 26 in the final sample) comparing the facial (surface) electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) concluded that \emph{self-reported} inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is hasty and based on waggly evidence. Indeed, concluding on the absence of an effect based on an under-powered non-significant \emph{p}-value is strongly problematic/uninformative. Moreover, the relation between self-reports and physiological measures was not \emph{directly} assessed, but only indirectly inferred from differences (or absence thereof) in group means. Given the ample inter-individual variability in these measures (as suggested by our reanalysis), we think inferring the individual-level relation between self-reports and physiological measures from group means is inappropriate. Given these limitations, we conclude that there is limited evidence for the main conclusion put forward by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) and we suggest ways forward, both from a theoretical and from a methodological perspective. Complete source code, reproducible analyses, and figures are available at \url{https://github.com/lnalborczyk/inner_experience_EMG}.
}

@@

Wordcount (excluding abstract, references, tables, and figures): 3934

@@

The main conclusion from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) is that inner experience between induced rumination and distraction differs ``without a change in electromyographic correlates of inner speech''. In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the OOI and OOS muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative for other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference has to be based on sufficiently powered sample size (given the target effect size) as well as on reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the present article. Moreover, a simple visual exploration of the data reveals important variability between individuals in the main effect of interest. That is, some participants had higher perioral (OOS and OOI) muscular activity in the rumination condition than in the distraction condition, and some other participants showed the reverse pattern. This suggests unexplored variation in the determinants of this effect (e.g., the content of the inner experience). Indeed, the relation between the inner experience and the physiological correlates of inner speech production was only inferred from group means. However, given the important inter-individual variability, this reasoning appears highly problematic. In the following, we explore each of these limitations and suggests ways forward, both from a theoretical and from a methodological perspective.

@@

As typical in studies manipulating induced rumination, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) designed a two-step protocol. First, they aimed to induce a negative mood by asking participants unsolvable and excessively difficult anagram and subtraction tasks, respectively. Second, they prompted the participants to either ruminate on these (purportedly induced) negative feelings (by asking them to ``think about the causes, consequences, and meaning of their current feelings'') or to distract themselves (by asking them to ``think about a village, city or town that you are particularly familiar with''). Rumination and distraction was manipulated within-subject, with all subjects alternating between rumination and distraction, in a counter-balanced order.

@@

As highlighted by several authors (e.g., Cohen, \protect\hyperlink{ref-cohen_earth_1994}{1994}; Pollard \& Richardson, \protect\hyperlink{ref-pollard_probability_1987}{1987}; Rouder et al., \protect\hyperlink{ref-rouder_is_2016}{2016}), concluding that an effect is probably absent solely based on a non-significant \emph{p}-value is the continuous (i.e., probabilistic) extension of the modus tollens and is not a valid argument (i.e., the conclusion does not follow from the premises). This fallacious argument is also known as the \emph{fallacy of acceptance}, the \emph{absence of evidence fallacy} or \emph{the argument from ignorance}, and proceeds as follows: "If the null hypothesis is true, then this observation should \emph{rarely} occur. This observation occurred. Therefore, the null hypothesis is false (or has low probability). In short, this argument is fallacious because it fails to consider the alternative hypothesis.

This problem is tackled in modern usages of null-hypothesis significance tests by ensuring that the claim under scrutiny is submitted to \emph{severe} tests (e.g., Mayo \& Spanos, \protect\hyperlink{ref-mayo_severe_2006}{2006}; Mayo, \protect\hyperlink{ref-mayo_statistical_2018}{2018}). In general terms, the strong severity principle states that we have evidence for a claim to the extent that it survives a stringent scrutiny, that is, to the extent that it survives \emph{severe} tests. More precisely, some claim (e.g., \(\theta = 0\)) is said to be \emph{severely tested} if it had great chances of being falsified, was the claim false. When a statistical test is under-powered (for detecting a given effect size) the claim under scrutiny is not strongly (severely) tested, hence it not possible to obtain strong or reliable evidence for the claim (bad test, no evidence).

@@

As shown in Figure \ref{fig:simulated-power}, the natural logarithm of the BF in favour of the alternative hypothesis is growing proportionally with sample size. More precisely, whereas BFs computed on small samples (i.e., below 80 participants) support the null hypothesis, BFs computed on adequately-powered samples support the alternative hypothesis for all three facial muscles. For instance, the average \(\text{BF}_{10}\) computed for the OOI muscle with a sample size of 160 participants is of \(\exp(2.18) \approx 8.85\), indicating that these data are approximately \(8.85\) times more likely under the alternative hypothesis than under the null hypothesis. To sum up, this reveals that although at low sample sizes, the BF may provide (weak) evidence for the null hypothesis (relative to the alternative hypothesis) this pattern may very well reverse for higher sample sizes.

We should keep in mind some limitations of this analysis, which uses simulated datasets form the posterior predictive distribution estimated on the data collected by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}). This analysis resembles to a loose Bayesian analogue of the frequentist post-hoc power analysis, which has been much criticised (e.g., Lakens, \protect\hyperlink{ref-lakens_20_2014}{2014}). A crucial assumption of the present analysis is that the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) is our best source of information regarding the main effect of interest. However, the present analysis also differs from the frequentist post-hoc power analysis on several grounds. First, with the present analysis, we do not aim to assess the ability of our statistical test to pass some dichotomic threshold (e.g., accept/reject). Instead, we aim to assess how the \(\text{BF}_{10}\) (i.e., the evidence for the alternative hypothesis, relative to the null hypothesis) behaves with varying sample sizes. Second, the present analysis relies on the posterior predictive distribution of the model fitted on the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), which naturally incorporates uncertainty about the effect of interest. By simulating datasets of varying sample sizes from the posterior predictive distribution (and by relying on a large number of simulations), uncertainty about the effect size is naturally incorporated into the simulation.

@@

We previously noted (e.g., Nalborczyk et al., \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}; Nalborczyk, \protect\hyperlink{ref-nalborczyk_understanding_2019}{2019}; Nalborczyk, Banjac, et al., \protect\hyperlink{ref-nalborczyk_dissociating_2020}{2020}; Nalborczyk, Grandchamp, et al., \protect\hyperlink{ref-nalborczyk_can_2020}{2020}) that surface EMG measures of inner speech production were highly variable between individuals. This can be explained by the imagery ability of each individual, the reliability of the measurement, and the instructions that are given to the participants (and whether they are understood in a similar manner by all participants). The data collected by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) is no exception and present an important degree of inter-individual variability. In Figure \ref{fig:everyone}, we represent again the EMG data for each participant (each line is a participant). We used two colours to represent the participants that showed a higher average EMG peak amplitude either in the rumination condition (in green) or in the distraction condition (in orange). As it can be seen from this figure, whereas some participants show ``intermediate'' or ``ambiguous'' patterns across the conditions, some participants show a clear superior EMG peak amplitude in the rumination condition (in green) or in the distraction condition (in orange).

@@

This important inter-individual variability calls into question the use of group averages to describe the nature of inner speech at an individual level. Moreover, this variability suggests that some important confounding factors were not taken into account (i.e., either not manipulated in the experiment or statistically controlled for). In line with Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), we suggest these discrepancies could be explained by differences in the subjective experience of inner speech. We agree that a lot could be learnt by relating this (self-reported) subjective experience to the peripheral muscular correlates of inner speech production. However, this can not be done at the group level, at the risk of missing individual-level patterns. Therefore, we encourage Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) to re-analyse their data to assess whether the perioral EMG correlates (e.g., the amplitude of the difference between the rumination and distraction conditions on the OOI) can be predicted by the self-reported subjective experience, \emph{at an individual level}.

It should be noted that the question of the qualitative differences in the EMG correlates of inner speech may also be assessed more formally using the model comparison approach developed by Haaf and Rouder (\protect\hyperlink{ref-haaf_developing_2017}{2017}). However, this would require data coming from an experimental design in which inner speech and non-inner speech conditions would be manipulated within-subject and with multiple observations for each participants in each condition (e.g., as in Nalborczyk, Grandchamp, et al., \protect\hyperlink{ref-nalborczyk_can_2020}{2020}), so that multilevel models with both varying intercepts and varying slopes could be estimated.

@@

In addition to these methodological limitations, we now wish the theoretical interpretations and implications of these results. As discussed in the introduction section, we previously conducted several studies aiming to assess the role of the speech motor system in rumination. Following our initial study (Nalborczyk et al., \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}), we ran an extension of this study in which we compared verbal to non-verbal rumination. The results of this study suggested that the facial EMG correlates of verbal and non-verbal rumination were similar (Nalborczyk, Banjac, et al., \protect\hyperlink{ref-nalborczyk_dissociating_2020}{2020}). Given the ample evidence on the EMG correlates of inner speech production (for an overview, see Chapter 1 in Nalborczyk, \protect\hyperlink{ref-nalborczyk_understanding_2019}{2019}), we needed to explain why this particular form of inner speech (induced rumination) was not associated with speech-specific peripheral muscular activity.

In Nalborczyk, Banjac, et al. (\protect\hyperlink{ref-nalborczyk_dissociating_2020}{2020}), we suggested that this observation was coherent with the mental-habit view of depressive rumination (Watkins \& Nolen-Hoeksema, \protect\hyperlink{ref-watkins_habit-goal_2014}{2014}), which defines rumination as a habitual behaviour, automatically triggered by contextual cues such as negative mood. Indeed, habitual behaviours are more automatic (they are not intentionally initiated) than non-habitual behaviours. It has been observed that the automaticity with which a verbal thought is evoked may influence the degree to which it is enacted, that is, the (e.g., Cohen, \protect\hyperlink{ref-cohen_motor_1986}{1986}; Sokolov, \protect\hyperlink{ref-sokolov_inner_1972}{1972})\ldots{}

@@

To conclude, we wish to bring some nuance to the conclusion of Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), who stated that ``In conclusion, induced rumination appeared to involve similar levels of inner speech-related muscle activity to a period of distraction''. In consideration of the limitations discussed in the present article, this conclusion seems hasty. Indeed, we provided theoretical (epistemological) and empirical (via simulation) reasons to doubt the strength of the evidence for the null hypothesis in this study. Moreover, supplementary analyses showed that the order of the conditions participants went through may have influenced the effects of the rumination induction on the EMG correlates. Finally, important under-explored inter-individual variability suggests that important determinants of these correlates were not taken into account. We urge the authors to nuance their conclusions, to reanalyse their data, and to plan adequately powered future studies in order to settle these issues.

>>>>

\abstract{
Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) reported the results of an experiment (N = 26 in the final sample) comparing the facial (surface) electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) concluded that \emph{self-reported} inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is at best hasty. Indeed, concluding on the absence of an effect based on an under-powered non-significant \emph{p}-value is strongly problematic/uninformative. Moreover, the relation between self-reports and physiological measures was not directly assessed, but only indirectly inferred from differences (or absence thereof) in group averages. Given the ample inter-individual variability in these measures (as suggested by our reanalysis), we think inferring the individual-level relation between self-reports and physiological measures from group averages is inappropriate. Given these limitations, we conclude that there is limited evidence for the main conclusion put forward by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) and we suggest ways forward, both from a theoretical and from a methodological perspective. Complete source code, reproducible analyses, and figures are available at \url{https://github.com/lnalborczyk/inner_experience_EMG}.
}

@@

Wordcount (excluding abstract, references, tables, and figures): 3933

@@

The main conclusion from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) is that inner experience between induced rumination and distraction differs ``without a change in electromyographic correlates of inner speech''. In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the OOI and OOS muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative for other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference has to be based on sufficiently powered sample size (given the target effect size) as well as on reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the present article. Moreover, a simple visual exploration of the data reveals important variability between individuals in the main effect of interest. That is, some participants had higher perioral (OOS and OOI) muscular activity in the rumination condition than in the distraction condition, and some other participants showed the reverse pattern. This suggests unexplored variation in the determinants of this effect (e.g., the content of the inner experience). Indeed, the relation between the inner experience and the physiological correlates of inner speech production was only inferred from group averages. However, given the important inter-individual variability, this reasoning appears highly problematic. In the following, we explore each of these limitations and suggests ways forward, both from a theoretical and from a methodological perspective.

@@

As typical in studies manipulating induced rumination, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) designed a two-step protocol. First, they aimed to induce a negative mood by asking participants to solve unsolvable or excessively difficult anagram and subtraction tasks. Second, they prompted participants to either ruminate on these (purportedly induced) negative feelings (by asking them to ``think about the causes, consequences, and meaning of their current feelings'') or to distract themselves (by asking them to ``think about a village, city or town that you are particularly familiar with''). Rumination and distraction was manipulated within-subject, with all subjects alternating between rumination and distraction, in a counter-balanced order.

@@

As highlighted by several authors (e.g., Cohen, \protect\hyperlink{ref-cohen_earth_1994}{1994}; Pollard \& Richardson, \protect\hyperlink{ref-pollard_probability_1987}{1987}; Rouder et al., \protect\hyperlink{ref-rouder_is_2016}{2016}), concluding that an effect is probably absent solely based on a non-significant \emph{p}-value is the continuous (i.e., probabilistic) extension of the modus tollens and is not a valid argument (i.e., the conclusion does not follow from the premises). This fallacious argument is also known as the \emph{fallacy of acceptance}, the \emph{absence of evidence fallacy} or \emph{the argument from ignorance}, and proceeds as follows: ``If the null hypothesis is true, then this observation should \emph{rarely} occur. This observation occurred. Therefore, the null hypothesis is false (or has low probability)''. In short, this argument is fallacious because it fails to consider the alternative hypothesis.

This problem is tackled in modern usages of null-hypothesis significance tests by ensuring that the claim under scrutiny is submitted to \emph{severe} tests (e.g., Mayo \& Spanos, \protect\hyperlink{ref-mayo_severe_2006}{2006}; Mayo, \protect\hyperlink{ref-mayo_statistical_2018}{2018}). In general terms, the strong severity principle states that we have evidence for a claim to the extent that it survives a stringent scrutiny, that is, to the extent that it survives \emph{severe} tests. More precisely, some claim (e.g., \(\theta = 0\)) is said to be \emph{severely tested} if it had great chances of being falsified, was the claim false. When a statistical test is under-powered (for detecting a given effect size), the claim under scrutiny is not strongly (severely) tested, hence it not possible to obtain strong or reliable evidence for the claim (bad test, no evidence).

@@

As shown in Figure \ref{fig:simulated-power}, the natural logarithm of the BF in favour of the alternative hypothesis is growing proportionally with sample size. More precisely, whereas BFs computed on small samples (i.e., below 80 participants) support the null hypothesis, BFs computed on adequately-powered samples support the alternative hypothesis for all three facial muscles. For instance, the average \(\text{BF}_{10}\) computed for the OOI muscle with a sample size of 160 participants is of \(\exp(2.18) \approx 8.85\), indicating that these data are approximately \(8.85\) times more likely under the alternative hypothesis than under the null hypothesis. To sum up, this reveals that although at low sample sizes, the BF may provide (weak) evidence for the null hypothesis (relative to the alternative hypothesis), this pattern may very well reverse for higher sample sizes.

We should keep in mind some limitations of this analysis, which uses simulated datasets form the posterior predictive distribution estimated on the data collected by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}). This analysis resembles to the Bayesian analogue of the frequentist post-hoc power analysis, which has been much criticised (e.g., Lakens, \protect\hyperlink{ref-lakens_20_2014}{2014}). A crucial assumption of the present analysis is that the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) is our best source of information regarding the main effect of interest. However, the present analysis also differs from the frequentist post-hoc power analysis on several grounds. First, with the present analysis, we do not aim to assess the ability of our statistical test to pass some dichotomic threshold (e.g., accept/reject). Instead, we aim to assess how the \(\text{BF}_{10}\) (i.e., the evidence for the alternative hypothesis, relative to the null hypothesis) behaves with varying sample sizes. Second, the present analysis relies on the posterior predictive distribution of the model fitted on the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), which naturally incorporates uncertainty about the effect of interest. By simulating datasets of varying sample sizes from the posterior predictive distribution (and by relying on a large number of simulations), uncertainty about the effect size is naturally incorporated into the results of the simulation.

@@

We previously noted (e.g., Nalborczyk et al., \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}; Nalborczyk, \protect\hyperlink{ref-nalborczyk_understanding_2019}{2019}; Nalborczyk, Banjac, et al., \protect\hyperlink{ref-nalborczyk_dissociating_2020}{2020}; Nalborczyk, Grandchamp, et al., \protect\hyperlink{ref-nalborczyk_can_2020}{2020}) that surface EMG measures of inner speech production were highly variable between individuals. This can be explained by the imagery ability of each individual, the reliability of the measurement, and the instructions that are given to the participants (and whether they are understood in a similar manner by all participants). The data collected by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) is no exception and present an important degree of inter-individual variability. In Figure \ref{fig:everyone}, we represent again the EMG data for each participant (each line is a participant). We used two colours to represent the participants that showed a higher average EMG peak amplitude either in the rumination condition (in green) or in the distraction condition (in orange). As it can be seen from this figure, whereas some participants show ``intermediate'' or ``ambiguous'' (i.e., equivalent) patterns of muscular activity across conditions, some participants show a clear superior EMG peak amplitude in the rumination condition (in green) and some others in the distraction condition (in orange).

@@

This important inter-individual variability calls into question the use of group averages to describe the nature of inner speech at an individual level. Moreover, this variability suggests that some important confounding factors were not taken into account (i.e., either not manipulated in the experiment or statistically controlled for). In line with Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), we suggest these discrepancies could be explained by differences in the subjective experience of inner speech. We agree that a lot could be learnt by relating this (self-reported) subjective experience to the peripheral muscular correlates of inner speech production. However, this can not be done at the group level, at the risk of missing individual-level patterns. Therefore, we encourage Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) to analyse further their data in order to assess whether the perioral EMG correlates (e.g., the amplitude of the difference between the rumination and distraction conditions on the OOI) can be predicted by the self-reported subjective experience, \emph{at an individual level}.

It should be noted that the question of the qualitative differences in the EMG correlates of inner speech may also be assessed more formally using the model comparison approach developed by Haaf and Rouder (\protect\hyperlink{ref-haaf_developing_2017}{2017}). However, this would require data coming from an experimental design in which inner speech and non-inner speech conditions would be manipulated within-subject and with multiple observations for each participants in each condition (e.g., as in Nalborczyk, Grandchamp, et al., \protect\hyperlink{ref-nalborczyk_can_2020}{2020}).

@@

In addition to these methodological limitations, we now wish to discuss the theoretical interpretations and implications of these results. As discussed in the introduction section, we previously conducted several studies aiming to assess the role of the speech motor system in rumination. Following our initial study (Nalborczyk et al., \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}), we ran an extension of this study in which we compared verbal to non-verbal rumination. The results of this study suggested that the facial EMG correlates of verbal and non-verbal rumination were similar (Nalborczyk, Banjac, et al., \protect\hyperlink{ref-nalborczyk_dissociating_2020}{2020}). Given the ample evidence on the EMG correlates of inner speech production (for an overview, see Chapter 1 in Nalborczyk, \protect\hyperlink{ref-nalborczyk_understanding_2019}{2019}), we needed to explain why this particular form of inner speech (induced rumination) was not associated with speech-specific peripheral muscular activity.

In Nalborczyk, Banjac, et al. (\protect\hyperlink{ref-nalborczyk_dissociating_2020}{2020}), we suggested that this observation was coherent with the mental-habit view of depressive rumination (Watkins \& Nolen-Hoeksema, \protect\hyperlink{ref-watkins_habit-goal_2014}{2014}), which defines rumination as a habitual behaviour, automatically triggered by contextual cues such as negative mood. We know habitual behaviours are more automatic (they are not intentionally initiated) than non-habitual behaviours. It has been observed that the automaticity with which a verbal thought is evoked may influence the degree to which it is enacted, that is, the (e.g., Cohen, \protect\hyperlink{ref-cohen_motor_1986}{1986}; Sokolov, \protect\hyperlink{ref-sokolov_inner_1972}{1972})\ldots{}

@@

To conclude, we wish to bring some nuance to the conclusion of Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), who stated that ``In conclusion, induced rumination appeared to involve similar levels of inner speech-related muscle activity to a period of distraction''. In consideration of the limitations discussed in the present article, this conclusion seems hasty. Indeed, we provided theoretical (epistemological) and empirical (via simulation) reasons to doubt the strength of the evidence for the null hypothesis in this study. Moreover, supplementary analyses showed that the order of the conditions participants went through may have influenced the effects of the rumination induction on the EMG correlates. Finally, important under-explored inter-individual variability suggests that important determinants of these correlates were not taken into account. We urge the authors to nuance their conclusions, to analyse further their data, and to plan adequately-powered studies in order to settle these issues.
